---
title: "Portfolio Optimization Using a Novel Data-Driven EWMA Covariance Model"
author: "Sulalitha Bowala"
date: "2023-06-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading Package
```{r}
library(zoo)
library(tseries)
library(corrplot)
library(xts)
library(xtable)
library(moments)
```


Load Data
```{r}
BTCdata <- read.csv('BTC_data_hourly.csv')
BTC=BTCdata$Adj.Close

ETHdata <- read.csv('ETH_data_hourly.csv')
ETH=ETHdata$Adj.Close

BNCdata <- read.csv('BNC_data_hourly.csv')
BNC=BNCdata$Adj.Close

XRPdata <- read.csv('XRP_data_hourly.csv')
XRP=XRPdata$Adj.Close

DOGEdata <- read.csv('DOGE_data_hourly.csv')
DOGE=DOGEdata$Adj.Close

ADAdata <- read.csv('ADA_data_hourly.csv')
ADA=ADAdata$Adj.Close

datecolumn = BTCdata$X

dataSet=cbind.data.frame(datecolumn, BTC, ETH, BNC, XRP, DOGE, ADA)
dim(dataSet)
dataSet=dataSet[-c(1:5040),] ## Remove first 7 months data
dim(dataSet)

head(dataSet)
```



```{r}
## calculating returns - entire dataset
#n_all = dim(data)[1]
#R_all <- (data[2:n_all, ] / data[1:(n_all-1), ] - 1)
dataSet1 = dataSet[-c(3471:11022),]

plot(dataSet1$BTC, type = "l", ylab='Adjusted Closing Price (USD)', main='Bitcoin', cex.main = 1.5, cex.lab = 1, cex.axis = 1, lwd=3.0, col = 1)

plot(dataSet1$ETH, type = "l", ylab='Adjusted Closing Price (USD)', main='Ethereum', cex.main = 1.5, cex.lab = 1, cex.axis = 1, lwd=3.0, col = "2")

plot(dataSet1$BNC, type = "l", ylab='Adjusted Closing Price (USD)', main='Binance Coin', cex.main = 1.5, cex.lab = 1, cex.axis = 1, lwd=3.0, col = 3)

plot(dataSet1$XRP, type = "l", ylab='Adjusted Closing Price (USD)', main='Ripple', cex.main = 1.5, cex.lab = 1, cex.axis = 1, lwd=3.0, col = 4)

plot(dataSet1$DOGE, type = "l", ylab='Adjusted Closing Price (USD)', main='Dogecoin', cex.main = 1.5, cex.lab = 1, cex.axis = 1, lwd=3.0, col = 5)

plot(dataSet1$ADA, type = "l", ylab='Adjusted Closing Price (USD)', main='Cardano', cex.main = 1.5, cex.lab = 1, cex.axis = 1, lwd=3.0, col = 6)
```


```{r}
### Selecting January data
dataSet = dataSet[-c(3468:11022),]
data<-as.matrix(dataSet[,-1])
colnames(data)<-c("Bitcoin", "Ethereum", "BinanceCoin", "Ripple", "Dogecoin", "Cardano")
tail(dataSet)
```



```{r}
prices<-tail(data, 721)
n = dim(prices)[1]
##### calculating returns for last 2800 values (testing data). i.e. last 20 months data
R =  (prices[2:n, ] / prices[1:(n-1), ] - 1)
number=ncol(data)
```


risk-free rate
```{r}
#risk-free rate for this study is the average treasury bill rate (T-bill rate) from April 1st, 2020, to February 18, 2022. All the data are obtained from Bloomberg.

Rf <- read.csv("Tbill_May.csv")
mufree <- mean(Rf[,2])/(365*100)
```


summary statistics
```{r}
apply(R, 2, mean)
apply(R, 2, sd)
apply(R, 2, skewness)
apply(R, 2, kurtosis)
```



calculate the lag 1 ACF of return, absolute value of return and square return
```{r}
acf.R<-matrix(0, nrow=number, ncol=3)
for(j in 1:number){
  acf.R[j, ]<-c(acf(R[, j], plot=FALSE)$acf[2],
                acf(abs(R[, j]), plot=FALSE)$acf[2],
                acf(R[, j]^2, plot=FALSE)$acf[2])
  ###### acf[2] means lag 1 sample auto-correlation
}
rownames(acf.R)<-colnames(R)
colnames(acf.R)<-c("series", "abs", "sq")
acf.R
```


```{r}
rho<-rep(0, number)
for(j in 1:number){
  rho[j]<-cor(R[, j]-mean(R[, j]), sign(R[, j]-mean(R[, j])))
}
names(rho)<-colnames(data)
rho
nu<-rep(0, number)
for(j in 1:number){
  fun <- function (x) rho[j]*(x-1)*beta(x/2,1/2)-2*sqrt(x-2)
  nu[j] <- uniroot(fun, c(2, 8))$root
}
nu
```


Empirical:
```{r}
mean_vect = apply(R,2,mean)
mean_vect

cov_mat<-cov(R)
cov_mat

z<-solve(cov_mat)%*%(mean_vect-mufree)
z

w.sigma.emp<-z/sum(z)
w.sigma.emp
```


Data Driven:
```{r}
#### choosing optimal alpha for EWMA
dd.alpha<-function(Z){
  alpha<-seq(0.01, 0.3, 0.01)
  t<-length(Z)
  cut.t<-200
  MSE_alpha<-rep(0, length(alpha))
  for(a in 1:length(alpha)){
    s<-mean(Z[1:cut.t])
    error<-rep(0, t)
    for(i in 1:t){
      error[i]<-Z[i]-s
      s<-alpha[a]*Z[i]+(1-alpha[a])*s
    }
    MSE_alpha[a]<-mean(error[-(1:cut.t)]^2)
  }
  alpha.opt<-alpha[which.min(MSE_alpha)]
  return(alpha.opt)
}
```



```{r}
#### data driven method to forecast volatility
dd.vol<-function(y){
  t<-length(y)
  
  rho<-cor(y-mean(y), sign(y-mean(y)))
  vol<-abs(y-mean(y))/rho
  #rho_vol <- cor(abs(y-mean(y)),(y-mean(y))^2)
  #vol <- sqrt(1-rho_vol^2)*sd(y)
    
  alpha<-dd.alpha(vol)
  cut.t<-500
  s<-mean(vol[1:cut.t])
  for(i in 1:t){
    s<-alpha*vol[i]+(1-alpha)*s
  }
  return(s)
}
```



```{r}
#### data driven approach to calculate residuals
dd.res<-function(y){
  
  rho<-cor(y-mean(y), sign(y-mean(y)))
  vol<-abs(y-mean(y))/rho
  #rho_vol <- cor(abs(y-mean(y)),(y-mean(y))^2)
  #vol <- sqrt(1-rho_vol^2)*sd(y)
  
  alpha<-dd.alpha(vol)
  cut.t<-500
  s<-mean(vol[1:cut.t])
  t<-length(y)
  res<-y
  for(i in 1:t){
    res[i]<-(y[i]-mean(y))/s
    s<-alpha*vol[i]+(1-alpha)*s
  }
  return(res)
}
```


```{r}
dd.cor.mat<-function(R){
  R.res<-R
  for(j in 1:ncol(R)){
    R.res[, j]<-dd.res(R[, j])
  }
  cor_mat<-cor(R.res)
  return(cor_mat)
}
```


```{r}
dd<-function(y){
  alpha<-seq(0.01, 0.3, 0.01)
  t<-length(y)
  cut.t<-500 
  
  rho<-cor(y-mean(y), sign(y-mean(y)))
  vol<-abs(y-mean(y))/rho
  #rho_vol <- cor(abs(y-mean(y)),(y-mean(y))^2)
  #vol <- sqrt(1-rho_vol^2)*sd(y)
  
  MSE_alpha<-rep(0, length(alpha))
  sn<-rep(0, length(alpha))
  for(a in 1:length(alpha)){
    s<-mean(vol[1:cut.t])
    error<-rep(0, t)
    for(i in 1:t){
      error[i]<-vol[i]-s
      s<-alpha[a]*vol[i]+(1-alpha[a])*s
    }
    MSE_alpha[a]<-mean(error[-(1:cut.t)]^2)
    sn[a]<-s
  }
  vol.fore<-sn[which.min(MSE_alpha)]
  return(vol.fore)
}
```


```{r}
mean_vect = apply(R,2,mean)
mean_vect

cor_mat<-dd.cor.mat(R)
cor_mat

sd_vect<-apply(R, 2, dd)
sd_vect

cov_mat<-sd_vect%*%t(sd_vect)*cor_mat
cov_mat

z<-solve(cov_mat)%*%(mean_vect-mufree)
z

w.sigma.dd<-z/sum(z)
w.sigma.dd

cor_mat.dd<-cor_mat

cov_mat.dd<-cov_mat

```


```{r}
portfolio_returns = function(w, asset_returns) {
  port.returns = 0
  
  # Multiplication of the i-th asset by the i-th weight in "x"
  for (i in 1:length(w)) {
    port.returns = port.returns + asset_returns[,i] * w[i]
  }
  
  return (port.returns)
}
```


```{r}
w.sigma<-data.frame(w.sigma.dd, w.sigma.emp, diag(1, number))
colnames(w.sigma)<-c("DD", "EMP", colnames(data))
w.sigma
sharpe.ratio<-c()
for(j in 1:ncol(w.sigma)){
  w<-w.sigma[, j]
  R.port<-portfolio_returns(w, R)
  vol<-dd(R.port)
  sharpe.ratio[j]<-(mean(R.port)-mufree)/vol
}
names(sharpe.ratio)<-c("DD", "EMP", colnames(data))
sharpe.ratio
```


```{r}
corrplot(cor(R), method="ellipse", main="Empirical")
corrplot(cor_mat.dd, method="ellipse", main="Data Driven")
cor(R); cor_mat.dd
cov(R); cov_mat.dd
```



Calculating sign-correlation and degree of freedom

```{r}
rho.cal<-function(R){
  rho<-cor(R-mean(R), sign(R-mean(R)))
  return(rho)
}
nu.cal<-function(R){
  rho<-cor(R-mean(R), sign(R-mean(R)))
  fun <- function (x) rho*(x-1)*beta(x/2,1/2)-2*sqrt(x-2)
  nu <- uniroot(fun, c(2, 9))$root
  return(nu)
}
```


calculate portfolio returns and sign-rho, and degree of freedom

```{r}
#### calculate the portfolio returns
R.port<-R%*%as.matrix(w.sigma)
rho<-apply(R.port, 2, rho.cal)
rho
nu<-apply(R.port, 2, nu.cal)
nu
```


VaR and ES calculation:
```{r}
#### forecasting volatility based on data driven method
vol.port<-apply(R.port, 2, dd)

#### calculating VaR
VaR.port<-(-1000)*qt(0.01, df=nu)*sqrt((nu-2)/nu)*vol
VaR.port

#### calculating ES
ES.cal<-function(nu, p) sqrt((nu-2)/nu)*dt(qt (p,nu), nu)/p*(nu + (qt (p,nu))^2)/(nu-1)
ES.port<-VaR.port
for(j in 1:number){
  ES.port[j]<-vol.port[j]*1000*ES.cal(nu[j], 0.01)
}
ES.port
```


summarize the portfolio and individual stock

```{r}
portfolio<-data.frame(sharpe.ratio,rho, nu, VaR.port, ES.port )
colnames(portfolio)<-c("Sharpe Ratio","rho", "df", "VaR", "ES")
portfolio
```


Sharpe ratio - empirical
```{r}
sharpe.emp.cal<-function(R, mufree){
  #mufree=0
  mean_vect = apply(R,2,mean)
  cov_mat<-cov(R)
  z<-solve(cov_mat)%*%(mean_vect-mufree)
  w<-z/sum(z)
  R.port<-portfolio_returns(w, R)
  vol<-dd(R.port)
  mad<-vol*cor(R.port-mean(R.port), sign(R.port-mean(R.port)))
  #sharpe.ma<-(mean(R.port)-mufree)/vol
  nu<-nu.cal(R.port)
  return(c(mean(R.port), vol, nu, mad))
}
```


Sharpe ratio - DD
```{r}
sharpe.dd.cal<-function(R, mufree){
  mean_vect = apply(R,2,mean)
  cor_mat<-dd.cor.mat(R)
  sd_vect<-apply(R, 2, dd)
  cov_mat<-sd_vect%*%t(sd_vect)*cor_mat
  #if(prod(eigen(cov_mat)$value>=0)==1) print("positive definite")
  #else print("negative definite")
  #mufree=0
  z<-solve(cov_mat)%*%(mean_vect-mufree)
  w<-z/sum(z)
  R.port<-portfolio_returns(w, R)
  vol<-dd(R.port)
  #sharpe.dd<-(mean(R.port)-mufree)/vol
  nu<-nu.cal(R.port)
  mad<-vol*cor(R.port-mean(R.port), sign(R.port-mean(R.port)))
  return(c(mean(R.port), vol, nu, mad))
}
```



```{r}
rownames(R) <- dataSet$datecolumn[-c(1:2747)]
R<-as.xts(R)
end.names<-rownames(as.matrix(R))
end.names<-tail(end.names, 24)
mufree<-Rf[, 2]
names(mufree)<-Rf$Date
mufree<-as.xts(mufree)
risk.free<-R[, 1]

temp = rep(mufree,each=26)
for (i in 1:length(risk.free)) {
  risk.free[i] = temp[i]
}

mufree<-as.vector(risk.free)/(100*365)
mufree<-tail(mufree, 24)
colnames(mufree)<-NULL
prices<-tail(data, 720+length(mufree))

n = dim(prices)[1]
R =  (prices[2:n, ] / prices[1:(n-1), ] - 1)

summary.emp<-matrix(0, nrow=length(mufree), ncol=4)
summary.dd<-matrix(0, nrow=length(mufree), ncol=4)
for(i in 1:length(mufree)){
  R.use<-as.data.frame(R[i:(720+i-1), ])
  summary.emp[i, ]<-sharpe.emp.cal(R.use, mufree[i])
  summary.dd[i, ]<-sharpe.dd.cal(R.use, mufree[i])
}

#head(summary.emp)
#head(summary.dd)

```


```{r}
mean.port<-data.frame(summary.emp[,1], summary.dd[, 1])
vol.port<-data.frame(summary.emp[, 2], summary.dd[, 2])
nu.port<-data.frame(summary.emp[, 3], summary.dd[, 3])
sharpe.port<-data.frame((mean.port[, 1]-mufree)/vol.port[, 1], 
                        (mean.port[, 2]-mufree)/vol.port[, 2])
mad.port<-data.frame(summary.emp[, 4], summary.dd[, 4])
colnames(mean.port)<-c("emp", "dd")
#rownames(mean.port)<-end.names
colnames(vol.port)<-c("emp", "dd")
#rownames(vol.port)<-end.names
colnames(nu.port)<-c("emp", "dd")
#rownames(nu.port)<-end.names
colnames(sharpe.port)<-c("emp", "dd")
#rownames(sharpe.port)<-end.names
colnames(mad.port)<-c("emp", "dd")
#rownames(mad.port)<-end.names
head(mean.port)
head(vol.port)
head(nu.port)
#### sd of mean of portfolio
mean_sd<-apply(mean.port, 2, sd)
head(mean_sd)
```


```{r}
VaR.port.dd<-(-1000)*(mean.port[, 2]+vol.port[, 2]*sqrt((nu.port[, 2]-2)/nu.port[, 2])*
                        qt(0.01, df=nu.port[, 2]))
VaR.port.emp<-(-1000)*(mean.port[, 1]+vol.port[, 1]*sqrt((nu.port[, 1]-2)/nu.port[, 1])*
                        qt(0.01, df=nu.port[, 1]))
ES.port.dd<-VaR.port.dd
ES.port.emp<-VaR.port.emp
for(i in 1:length(VaR.port.dd)){
  ES.port.dd[i]<-ES.cal(nu.port[i, 2], 0.01)*1000*vol.port[i, 2]
  ES.port.emp[i]<-ES.cal(nu.port[i, 1], 0.01)*1000*vol.port[i, 1]
}
VaR.port<-data.frame(VaR.port.emp, VaR.port.dd)
ES.port<-data.frame(ES.port.emp, ES.port.dd)
#rownames(VaR.port)<-end.names
#rownames(ES.port)<-end.names
colnames(VaR.port)<-c("emp", "dd")
colnames(ES.port)<-c("emp", "dd")
head(VaR.port)
head(ES.port)
```



```{r}
## summary - Empirical
vol.summary = c(mean(vol.port$emp), sd(vol.port$emp))
vol.summary

SR.summary = c(mean(sharpe.port$emp), sd(sharpe.port$emp))
SR.summary

MAD.summary = c(mean(mad.port$emp), sd(mad.port$emp))
MAD.summary

VaR.summary = c(mean(VaR.port$emp), sd(VaR.port$emp))
VaR.summary

ES.summary = c(mean(ES.port$emp), sd(ES.port$emp))
ES.summary
```



```{r}
## summary - Empirical
vol.summary = c(mean(vol.port$dd), sd(vol.port$dd))
vol.summary

SR.summary = c(mean(sharpe.port$dd), sd(sharpe.port$dd))
SR.summary

MAD.summary = c(mean(mad.port$dd), sd(mad.port$dd))
MAD.summary

VaR.summary = c(mean(VaR.port$dd), sd(VaR.port$dd))
VaR.summary

ES.summary = c(mean(ES.port$dd), sd(ES.port$dd))
ES.summary
```



```{r}
hist(sharpe.port[, 1], main="EMPIRICAL")
hist(sharpe.port[, 2], main="Data Driven")
```


```{r}
#main="Sharpe ratio Forecast", ylim=c(min(sharpe.port), max(sharpe.port)),
plot(sharpe.port[,1], type="S",lwd = 2, col="orange", ylab = "Sharpe ratio", ylim=c(min(sharpe.port), max(sharpe.port)))
lines(sharpe.port[,2],lwd = 2, col = "purple")
grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)
legend("topright", legend = c("EMP", "DD"), lty=c(1, 1), lwd=c(4, 4), col=c("orange", "purple"))

```


```{r}
#main="VaR Forecast",ylim=c(min(VaR.port), max(VaR.port)+2),
plot(VaR.port[,1], type="S", lwd = 2, col="orange", ylab = "Value-at-Risk", ylim = c(min(VaR.port), max(VaR.port)+2))
lines(VaR.port[,2], lwd = 2, col="purple")
grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)
legend("topright", legend = c("EMP", "DD"), lty=c(1, 1), lwd=c(4, 4), col=c("orange", "purple"))
#addLegend("topright", on=1, legend.names = c("EMP", "DD"), lty=c(1, 1), lwd=c(3, 3),col=c("blue", "red"))
```


```{r}
#main="ES Forecasts",ylim=c(min(ES.port), max(ES.port)),
plot(ES.port[,1], type="S", lwd = 2, col="orange", ylab = "Expected Shortfall", ylim = c(min(ES.port), max(ES.port)))
lines(ES.port[,2], lwd = 2, col = "purple")
grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)
legend("topright", legend = c("EMP", "DD"), lty=c(1, 1), lwd=c(4, 4), col=c("orange", "purple"))
#addLegend("topright", on=1, legend.names = c("EMP", "DD"), lty=c(1, 1), lwd=c(1, 1),col=c("blue", "red"))
```


```{r}
#main="Volatility Forecasts",ylim=c(min(vol.port), max(vol.port)+0.001),
plot(vol.port[,1], type="S", lwd = 2, col="orange", ylab = "Volatility", ylim = c(min(vol.port), max(vol.port)))
lines(vol.port[,2], lwd = 2, col = "purple")
grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)
legend("topright", legend = c("EMP", "DD"), lty=c(1, 1), lwd=c(4, 4), col=c("orange", "purple"))
#addLegend("topright", on=1, legend.names = c("EMP", "DD"), lty=c(1, 1), lwd=c(3, 3),col=c("blue", "red"))
```



```{r}
#main="MAD Forecasts",ylim=c(min(mad.port), max(mad.port)+0.001),
plot(mad.port[,1], type="S", lwd = 2, col="orange", ylab = "Mean Absolute Deviation", ylim = c(min(mad.port), max(mad.port)))
lines(mad.port[,2], lwd = 2, col = "purple")
grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)
legend("topright", legend = c("EMP", "DD"), lty=c(1, 1), lwd=c(4, 4), col=c("orange", "purple"))
#addLegend("topright", on=1, legend.names = c("EMP", "DD"), lty=c(1, 1), lwd=c(3, 3),col=c("blue", "red"))
```



```{r}
sharpe.ratio.mean<-apply(sharpe.port, 2, mean)
sharpe.ratio.sd<-apply(sharpe.port, 2, sd)
VaR.mean<-apply(VaR.port, 2, mean)
VaR.sd<-apply(VaR.port, 2, sd)
ES.mean<-apply(ES.port, 2, mean)
ES.sd<-apply(ES.port, 2, sd)
mad.sd<-apply(mad.port, 2, sd)
vol.sd<-apply(vol.port, 2, sd)
sharpe.ratio.mean; sharpe.ratio.sd
sharpe.summary<-data.frame(sharpe.ratio.mean, sharpe.ratio.sd)
sharpe.summary
                          
#mad.sd,VaR.sd,ES.sd)
#xtable(sharpe.summary, digits=4)

risk.summary<-data.frame(vol.sd, mad.sd,VaR.sd,ES.sd)
risk.summary
#xtable(risk.summary, digits=4)
```



```{r}
utility<-(sharpe.port[, 2]^2-sharpe.port[, 1]^2)/(sharpe.port[, 1]^2)*100
mean(utility)
utility<-as.data.frame(utility)
#colnames(utility)<-"% change in utility"
library(ggplot2)
ggplot(utility, aes(x=utility))+
  geom_histogram(color="black", fill="mistyrose1", binwidth = 20)+
  geom_vline(aes(xintercept=mean(utility)),
             color="navy", linetype="dashed", size=1)+
  labs(x="Percentage Change in Utility")
#title="Percentage Change in Utility - Histogram plot",
```


```{r}
library(ggplot2)
Approach<-c(rep("EMP", nrow(vol.port)),
             rep("DD", nrow(vol.port)))
data.plot<-data.frame(MAD=c(mad.port[, 1], mad.port[, 2]),
                      Volatility=c(vol.port[, 1], vol.port[, 2]),
                      VaR=c(VaR.port[, 1], VaR.port[, 2]), 
                      ES=c(ES.port[, 1], ES.port[, 2]),
                      Approach)
par(mfrow=c(4, 1))
ggplot(data.plot, aes(x=Approach, y=MAD, fill=Approach))+
  geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=2, notch=TRUE)
ggplot(data.plot, aes(x=Approach, y=Volatility, fill=Approach))+
  geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=2, notch=TRUE)
ggplot(data.plot, aes(x=Approach, y=VaR, fill=Approach))+
  geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=2, notch=TRUE)
ggplot(data.plot, aes(x=Approach, y=ES, fill=Approach))+
  geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=2, notch=TRUE)
#vol.p
```



```{r}
## Updated Figures

model.risk.E<-list()

model.risk.E[[1]]<-data.frame(vol.port[, 1], vol.port[, 2])
model.risk.E[[2]]<-data.frame(mad.port[, 1], mad.port[, 2])
model.risk.E[[3]]<-data.frame(VaR.port[, 1], VaR.port[, 2])
model.risk.E[[4]]<-data.frame(ES.port[, 1], ES.port[, 2])

main.names<-c("Volatility", "MAD", "VaR", "ES")

par(mfrow=c(1, 4))

for(j in 1:4){

  colnames(model.risk.E[[j]])<-c("EMP", "DD")

  boxplot(model.risk.E[[j]], col=c("yellow", 4), main=main.names[j])

}
```


